Haploid Selection, Sex Ratio Bias, and Transitions Between Sex-Determination Systems

### 2018-03-28 LaTeX/main_revision.tex
finished editing intro, added more complete discussion of Kozielska and Ubeda

%Haploid selection model feature and discussion of why needed
Furthermore, we include haploid selection in models describing cis-GSD, trans-GSD, and GSD to ESD transitions. 
Haploid selection (gametic competition or meiotic drive) is typically sex-specific in that it usually occurs among gametes produced by one sex only~\cite{Mulcahy:1996ha,JOSEPH:2004haa,Ubeda:2005gw,Lindholm:2016cw}. 
Therefore, one might expect new sex-determining systems to benefit from close linkage with haploid selected loci, as found for sex-specific diploid selection~\cite{vanDoorn:2007eu,vanDoorn:2010hu, Muralidhar2018}. 
On the other hand, associations between sex-determining loci and haploid selected loci generate biased zygotic sex ratios. 
Nevertheless, numerical simulations by Ubeda et al. (2015)~\cite{Ubeda:2015fx} suggest that, starting from GSD, new masculinizing mutations (neo-Y alleles) spread via associations with alleles that are beneficial in the male haploid stage (e.g., \cite{Ubeda:2015fx}), leading to male-biased sex ratios.

But Kozielska uses numerical simulations to show that ancestral sex ratio biases caused by haploid selection...

@@

%HOLES IN CURRENT THEORY
%(2) haploid selection; it can complicate outcomes for SD transitions
There are various ways by which haploid selection could influence transitions between sex-determining systems. 
If we assume that haploid selection at any particular locus predominantly occurs in one sex (e.g., meiotic drive during spermatogenesis), then such loci experience a form of sex-specific selection. 
In this respect, we might expect that haploid selection would affect transitions between sex-determining systems in a similar manner to sex-specific diploid selection (as explored in~\cite{vanDoorn:2007eu,vanDoorn:2010hu}). 
That is, new masculinizing mutations (neo-Y alleles) could be favoured via associations with alleles that are beneficial in the male haploid stage (e.g., \cite{Ubeda:2015fx}). 
Kozielska et al. (2010)~\cite{Kozielska:2010vm} have shown this to be theoretically possible for the special case of perfect linkage between the new sex-determining locus and a locus that experiences meiotic drive in one sex, under a limited number of forms of diploid selection.
Similarly, Ubeda et al. (2015)~\cite{Ubeda:2015fx} numerically show that sex-specific meiotic drive can allow masculinizing and feminizing mutations to spread in an ancestrally environmental-sex-determination system, again under a limited number of forms of diploid selection. 
On the other hand, sex ratios can also become biased by linkage between the sex-determining locus and a locus that harbours genetic variation in haploid fitness, as indeed occurs in the two examples above.
There are also several known empirical cases of sex-ratio bias caused by sex-linked meiotic drive alleles (Burt and Trivers, 2006, Chapter 6~\cite{Burt:2006}) or selection among X- and Y-bearing pollen~\cite{Lloyd:1974tz,Conn:1981uw,Stehlik:2005ul,Stehlik:2006to,Field:2012fd,Field:2013cc}. 
Sex-ratio selection can thus select against the spread of linked sex-determining alleles and favour the invasion of alternative, unlinked sex-determination systems \cite{Kozielska:2010vm,Ubeda:2015fx}.
%It is not immediately clear how the spread of new sex-determining systems would be influenced by the combination of sex-ratio biases and associations with haploid selected alleles. 
It is not yet clear when new sex-determining alleles are favoured by associations with haploid selected alleles vs. sex-ratio bias, especially under a wide range of possible diploid selection regimes (e.g., sex-antagonism, overdominance).

We consider both selected loci that are loosely linked to the sex-determining region (e.g., on an autosome) and selected loci that are in very tight linkage with the ancestral sex-determining locus (e.g., within the non-recombining sex-determining region).
Our main findings are three:
(1) transitions between sex-determining systems can decrease or increase sex-ratio bias, and with sufficiently loose linkage between the selected locus and the sex-determining locus, the selective force exerted by sex-ratio selection and favourable associations with haploid-selected alleles are exactly equivalent;
(2) ancestrally-tight sex-linkage and haploid selection allow trans-GSD transitions that weaken sex-linkage; and
(3) haploid selection removes the need for sex-antagonism, increasing the range of scenarios that select for transitions between sex-determining systems.
%MAIN FINDINGS IN GENERAL TERMS
%We find that the spread of novel sex-determining systems is influenced by both Fisherian sex-ratio selection and by selection on linked alleles. 
%We find that sex-ratio biases caused by haploid selection can exert Fisherian sex-ratio selection upon novel sex-determiners but that their spread is also determined by selection on genetically-associated alleles. 
%Indeed, it is only when haploid-selected loci are tightly linked to the ancestral sex-determining region (and so sex-ratio biases are initially large) that we see an asymmetry between selection for XY to ZW transitions and ZW to XY transitions (e.g., because haploid selection in males only causes biased zygotic sex ratios in an ancestrally XY system).
%Fisherian sex-ratio selection does not always dominate; it is possible for selection on linked alleles to drive turnover between sex-determining systems despite causing increasingly biased sex ratios, and our analytical results elucidate what is required for this to be so.
%Because sex-determining loci are often found within a region of suppressed recombination, there can be a significant number of tightly linked loci. 
%We find that loci tightly linked with the sex-determining locus can drive transitions in which the heterogametic sex changes, even when the new sex-determining locus is less closely linked to loci under selection (either including haploid selection or not).

@@

@@

%Limitation 3: don't consider m alleles with similar k to ancester, no ancestral ESD considered
In this study, we have only considered new sex-determining alleles of large effect. 
However, we expect similar selective forces to act on masculinizing/feminizing alleles of weaker effect.
For example, small effect masculinizing/feminizing alleles within a threshold model of sex determination can be favoured when linked to loci that experience sexually-antagonistic selection~\cite{Muralidhar2018}. 
These results echo those for large-effect neo-Y/neo-W alleles~\cite{vanDoorn:2007eu,vanDoorn:2010hu}.
\textcolor{red}{It should be noted, however, that the dynamics of sex-determining alleles with very weak effect will be influenced by genetic drift, which itself has been shown to cause trans-GSD transitions \cite{Veller2017}}.
Finally, while we have considered cis-GSD, trans-GSD, and GSD to ESD transitions, we have not explicitly considered ESD to GSD transitions. 
Recent models of ESD to GSD transitions~\cite{Ubeda:2015fx,Muralidhar2018} show that that neo-Y/neo-W alleles can be favoured when they arise near to haploid and/or diploid selected loci, which also occurs in our model.

>>>>

%Haploid selection model feature and discussion of why needed
Most significantly, we include haploid selection (gametic competition or meiotic drive) in models describing cis-GSD, trans-GSD, and GSD to ESD transitions, which poses an apparent evolutionary problem. 
On one hand, haploid selection is typically sex-specific in that it usually occurs among gametes produced by one sex only~\cite{Mulcahy:1996ha,JOSEPH:2004haa,Ubeda:2005gw,Lindholm:2016cw}. 
Therefore, one might expect new sex-determining systems to benefit from close linkage with haploid selected loci, as found for sex-differences in diploid selection~\cite{vanDoorn:2007eu,vanDoorn:2010hu, Muralidhar2018}. 
On the other hand, associations between sex-determining loci and haploid selected loci generate biased zygotic sex ratios, which should generally hinder the spread of new sex-determining systems.

%Previous models with haploid selection
Two previous numerical simulation studies have considered the spread of GSD with sex-specific meiotic drive~\cite{Kozielska:2010vm,Ubeda:2015fx} under a small number of possible genetic architectures and diploid selective regimes. 
Ubeda et al. (2015)~\cite{Ubeda:2015fx}, considers ancestral-ESD (with no sex ratio bias) and find that new GSD alleles can spread if they arise in linkage with meiotic drive loci. 
For example, a masculinizing allele spreads in association with an allele that is favoured during male meiosis, causing sex ratios to become male-biased and thereby suggesting that such beneficial associations overwhelm selection to balance the sex ratio. 
However, Kozielska et al. (2010)~\cite{Kozielska:2010vm} consider meiotic drive on an ancestral GSD sex chromosome (creating a sex ratio bias). 
They find that new, unlinked, GSD systems can spread if they generate the rarer sex, creating a balanced sex ratio. 
That is, they conclude that Fisherian sex ratio selection caused by haploid selection favours new GSD systems. 
Here, we analytically find conditions under which new GSD or ESD systems spread with generic linkage between the loci involved (and generic sex-specific haploid and diploid selection).

%main results
Strikingly, we reconcile the results of Kozielska et al. (2010)~\cite{Kozielska:2010vm} and Ubeda et al. (2015)~\cite{Ubeda:2015fx} by showing that new GSD systems commonly spread when they increase \textit{or} decrease linkage with loci that experience haploid selection.
This result is qualitatively distinct from those for diploid selection alone~\cite{vanDoorn:2007eu,vanDoorn:2010hu} and suggests that haploid selection is more likely to promote transitions between sex-determination systems. 
However, we also show that transitions involving haploid selection cannot be simply explained by invoking sex ratio selection. 
In particular, under a wide range of conditions, we show that haploid selection drives transitions in sex-determining system \textit{equally strongly} whether sex-ratio biases increase or decrease (and whether sex-ratio biases are present or absent). 
Finally, we show that ESD may not evolve, even if the sex-ratio is initially biased by haploid selection, which would not be predicted by previous theories for transitions to ESD~\cite{Charnov:1982wg,Bull:1983vi,West:2009we}.

@@

%HOLES IN CURRENT THEORY
%(2) haploid selection; it can complicate outcomes for SD transitions
%There are various ways by which haploid selection could influence transitions between sex-determining systems. 
%If we assume that haploid selection at any particular locus predominantly occurs in one sex (e.g., meiotic drive during spermatogenesis), then such loci experience a form of sex-specific selection. 
%In this respect, we might expect that haploid selection would affect transitions between sex-determining systems in a similar manner to sex-specific diploid selection (as explored in~\cite{vanDoorn:2007eu,vanDoorn:2010hu}). 
%That is, new masculinizing mutations (neo-Y alleles) could be favoured via associations with alleles that are beneficial in the male haploid stage (e.g., \cite{Ubeda:2015fx}). 
%Kozielska et al. (2010)~\cite{Kozielska:2010vm} have shown this to be theoretically possible for the special case of perfect linkage between the new sex-determining locus and a locus that experiences meiotic drive in one sex, under a limited number of forms of diploid selection.
%Similarly, Ubeda et al. (2015)~\cite{Ubeda:2015fx} numerically show that sex-specific meiotic drive can allow masculinizing and feminizing mutations to spread in an ancestrally environmental-sex-determination system, again under a limited number of forms of diploid selection. 
%On the other hand, sex ratios can also become biased by linkage between the sex-determining locus and a locus that harbours genetic variation in haploid fitness, as indeed occurs in the two examples above.
%There are also several known empirical cases of sex-ratio bias caused by sex-linked meiotic drive alleles (Burt and Trivers, 2006, Chapter 6~\cite{Burt:2006}) or selection among X- and Y-bearing pollen~\cite{Lloyd:1974tz,Conn:1981uw,Stehlik:2005ul,Stehlik:2006to,Field:2012fd,Field:2013cc}. 
%Sex-ratio selection can thus select against the spread of linked sex-determining alleles and favour the invasion of alternative, unlinked sex-determination systems \cite{Kozielska:2010vm,Ubeda:2015fx}.
%It is not yet clear when new sex-determining alleles are favoured by associations with haploid selected alleles vs. sex-ratio bias, especially under a wide range of possible diploid selection regimes (e.g., sex-antagonism, overdominance).

%We consider both selected loci that are loosely linked to the sex-determining region (e.g., on an autosome) and selected loci that are in very tight linkage with the ancestral sex-determining locus (e.g., within the non-recombining sex-determining region).
%Our main findings are three:
%(1) transitions between sex-determining systems can decrease or increase sex-ratio bias, and with sufficiently loose linkage between the selected locus and the sex-determining locus, the selective force exerted by sex-ratio selection and favourable associations with haploid-selected alleles are exactly equivalent;
%(2) ancestrally-tight sex-linkage and haploid selection allow trans-GSD transitions that weaken sex-linkage; and
%(3) haploid selection removes the need for sex-antagonism, increasing the range of scenarios that select for transitions between sex-determining systems.

%MAIN FINDINGS IN GENERAL TERMS
%We find that the spread of novel sex-determining systems is influenced by both Fisherian sex-ratio selection and by selection on linked alleles. 
%We find that sex-ratio biases caused by haploid selection can exert Fisherian sex-ratio selection upon novel sex-determiners but that their spread is also determined by selection on genetically-associated alleles. 
%Indeed, it is only when haploid-selected loci are tightly linked to the ancestral sex-determining region (and so sex-ratio biases are initially large) that we see an asymmetry between selection for XY to ZW transitions and ZW to XY transitions (e.g., because haploid selection in males only causes biased zygotic sex ratios in an ancestrally XY system).
%Fisherian sex-ratio selection does not always dominate; it is possible for selection on linked alleles to drive turnover between sex-determining systems despite causing increasingly biased sex ratios, and our analytical results elucidate what is required for this to be so.
%Because sex-determining loci are often found within a region of suppressed recombination, there can be a significant number of tightly linked loci. 
%We find that loci tightly linked with the sex-determining locus can drive transitions in which the heterogametic sex changes, even when the new sex-determining locus is less closely linked to loci under selection (either including haploid selection or not).

@@

\textcolor{red}{Discussion of `epistatic dominance' assumption to go here somewhere? The model in the Mathematica file is actual general to this assumption I think (or at least we wrote one that is), which we could mention in the response.}

@@

%Limitation 3: don't consider m alleles with similar k to ancester, no ancestral ESD considered
In this study, we have only considered new sex-determining alleles of large effect. 
However, we expect similar selective forces to act on masculinizing/feminizing alleles of weaker effect.
For example, small effect masculinizing/feminizing alleles within a threshold model of sex determination can be favoured when linked to loci that experience sexually-antagonistic selection~\cite{Muralidhar2018}. 
These results echo those for large-effect neo-Y/neo-W alleles~\cite{vanDoorn:2007eu,vanDoorn:2010hu}.
\textcolor{red}{It should be noted, however, that the dynamics of sex-determining alleles with very weak effect will be influenced by genetic drift, which itself has been shown to cause trans-GSD transitions \cite{Veller2017}}.
